\documentclass[11pt]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsfonts, amssymb, amsmath, amsthm, booktabs, hyperref, pgfplots, tikz, xcolor}

\theoremstyle{definition}\newtheorem{definition}{Definition}
\theoremstyle{definition}\newtheorem{notation}{Notation}
\theoremstyle{definition}\newtheorem{example}{Example}
\theoremstyle{definition}\newtheorem{question}{Question}
\theoremstyle{theorem}\newtheorem{theorem}{Theorem}
\theoremstyle{theorem}\newtheorem{corollary}{Corollary}
\theoremstyle{theorem}\newtheorem{proposition}{Proposition}
\theoremstyle{theorem}\newtheorem{lemma}{Lemma}
\theoremstyle{remark}\newtheorem{remark}{Remark}

\newcommand{\K}{\mathbb{K}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\CC}{\mathcal{C}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\F}{\mathbb{F}}
\renewcommand{\SS}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\teq}{\trianglelefteq}
\DeclareMathOperator{\diag}{diag}

\everymath{\displaystyle}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\begin{document}

\textbf{MATH 4271 Dynamical Systems} \hfill \textbf{Problem Set 1} \\
\textsc{Practice} \hfill \textsc{Joe Tran}

\begin{question}
    Find the general solution of the following linear systems
    \begin{itemize}
        \item[(a)] $\frac{dx}{dt} = x$, $\frac{dy}{dt} = y$
        \item[(b)] $\frac{dx}{dt} = x$, $\frac{dy}{dt} = 2y$
        \item[(c)] $\frac{dx}{dt} = x$, $\frac{dy}{dt} = 3y$
        \item[(d)] $\frac{dx}{dt} = -y$, $\frac{dy}{dt} = x$
        \item[(e)] $\frac{dx}{dt} = -x + y$, $\frac{dy}{dt} = -y$
    \end{itemize}
\end{question}

\begin{question}
    Find the general solution of the following linear systems.
    \begin{itemize}
        \item[(a)] $\frac{dx}{dt} = x$, $\frac{dy}{dt} = y$, $\frac{dz}{dt} = z$
        \item[(b)] $\frac{dx}{dt} = -x$, $\frac{dy}{dt} = -y$, $\frac{dz}{dt} = z$
        \item[(c)] $\frac{dx}{dt} = -y$, $\frac{dy}{dt} = x$, $\frac{dz}{dt} = -z$
    \end{itemize}
\end{question}

\begin{question}
    Find the general solution of the linear system
    \begin{equation*}
        \frac{dx}{dt} = x \quad \frac{dy}{dt} = ay
    \end{equation*}
    where $a$ is a constant.
\end{question}

\begin{question}
    Find the general solution of the linear system
    \begin{equation*}
        \frac{d\vec{x}}{dt} = A\vec{x} \tag{1}
    \end{equation*}
    when $A$ is the $n \times n$ diagonal matrix $A = \diag(\lambda_1, \lambda_2,..., \lambda_n)$. What condition on the eigenvalues $\lambda_1,..., \lambda_n$ will guarantee that $\lim_{t \to \infty} \vec{x}(t) = \vec{0}$ for all solutions $\vec{x}(t)$ of (1)?
\end{question}

\begin{question}
    What is the relationship between the vector fields defined by
    \begin{equation*}
        \frac{d\vec{x}}{dt} = A\vec{x}
    \end{equation*}
    and
    \begin{equation*}
        \frac{d\vec{x}}{dt} = kA\vec{x}
    \end{equation*}
    where $k$ is a nonzero constant?
\end{question}

\begin{question}
    \begin{itemize}
        \item[(a)] If $\vec{u}(t)$ and $\vec{v}(t)$ are solutions of the linear system (1), prove that for any constants $a$ and $b$, $\vec{w}(t) = a\vec{u}(t) + b\vec{v}(t)$ is a solution.
        \item[(b)] For
        \begin{equation*}
            A = \begin{bmatrix}
                1 & 0 \\
                0 & -2
            \end{bmatrix}
        \end{equation*}
        find solutions $\vec{u}(t)$ and $\vec{v}(t)$ of $\frac{d\vec{x}}{dt} = A\vec{x}$ such that every solution is a linear combination of $\vec{u}(t)$ and $\vec{v}(t)$.
    \end{itemize}
\end{question}

\begin{question}
    Find the eigenvalues and eigenvectors of the matrix $A$, and show that $D = P^{-1}AP$ is a diagonal matrix. Solve the linear system $\frac{d\vec{y}}{dt} = D\vec{y}$ and then solve $\frac{d\vec{x}}{dt} = A\vec{x}$.
    \begin{itemize}
        \item[(a)] $A = \begin{bmatrix} 1 & 3 \\ 3 & 1 \end{bmatrix}$
        \item[(b)] $A = \begin{bmatrix} 3 & 1 \\ 1 & 3 \end{bmatrix}$
        \item[(c)] $A = \begin{bmatrix} -1 & 1 \\ 1 & -1 \end{bmatrix}$
    \end{itemize}
\end{question}

\begin{question}
    Find the eigenvalues and eigenvectors for the matrix $A$, solve the linear system $\frac{d\vec{x}}{dt} = A\vec{x}$, determine the stable and unstable subspaces for the linear system
    \begin{equation*}
        \frac{d\vec{x}}{dt} = \begin{bmatrix} 1 & 0 & 0 \\ 1 & 2 & 0 \\ 1 & 0 & -1 \end{bmatrix}\vec{x}
    \end{equation*}
\end{question}

\begin{question}
    Write the following linear differential equations with constant coefficients in the form of the linear system (1), then solve.
    \begin{itemize}
        \item[(a)] $\frac{d^2x}{dt^2} + \frac{dx}{dt} - 2x = 0$
        \item[(b)] $\frac{d^2x}{dt^2} + x = 0$
        \item[(c)] $\frac{d^3x}{dt^2} - 2\frac{d^2x}{dt^2} - \frac{dx}{dt} + 2x = 0$
    \end{itemize}
\end{question}

\begin{question}
    Solve the initial value problem
    \begin{equation*}
        \frac{d\vec{x}}{dt} = A\vec{x} \quad \vec{x}(0) = \vec{x}_0
    \end{equation*}
    \begin{itemize}
        \item[(a)] with $A$ given by Question 7(a) above and $\vec{x}_0 = (1, 2)^T$
        \item[(b)] with $A$ given by Question 8 and $\vec{x}_0 = (1, 2, 3)^T$.
    \end{itemize}
\end{question}

\begin{question}
    Let $A$ be an $n \times n$ matrix with real and distinct eigenvalues. Find conditions on the eigenvalues that are necessary and sufficient for $\lim_{t \to \infty} \vec{x}(t) = \vec{0}$, where $\vec{x}(t)$ is any solution of $\frac{d\vec{x}}{dt} = A\vec{x}$.
\end{question}

\begin{question}
    Let $A$ be an $n \times n$ matrix with real and distinct eigenvalues. Let $\phi(t, \vec{x}_0)$ be the solution of the initial value problem
    \begin{equation*}
        \frac{d\vec{x}}{dt} = A\vec{x} \quad \vec{x}(0) = \vec{x}_0
    \end{equation*}
    Show that for each fixed $t \in \R$,
    \begin{equation*}
        \lim_{\vec{y}_0 \to \vec{x}_0} = \phi(t, \vec{y}_0) = \phi(t, \vec{x}_0)
    \end{equation*}
\end{question}

\end{document}